---
title: "Power simulation for example 1"
author: "Jim Matthews"
date: "18/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Chicken Welfare - bone density (difference between 2 means)  

load the required packages

```{r load packages, echo=TRUE, message=FALSE, warning=FALSE}
library('paramtest')
library('pwr')
library('tidyverse')
```

### Create a function to simulate data and perform t-test

the function creates two normally distributed samples of data, namely x1 and x2  
x1 has mean = 0 and SD = 1  
x2 has mean = d and SD = 1  
the function requires you to specify the standardised effect size, d  
and the sample size for each group, N  

```{r t_func, echo=TRUE}
t_func <- function(simNum, N, d) {
  x1 <- rnorm(N, 0, 1)
  x2 <- rnorm(N, d, 1)
  
  t <- t.test(x1, x2, var.equal = TRUE)  # run t-test
  stat <- t$statistic
  p <- t$p.value
  
  return(c(t = stat, p = p, sig = (p < .05)))
  # return a named vector with the results we want to keep
  # value of sig will be TRUE or FALSE based on the p value
}
```

### Specify the values of sample size N (per group) and cohens d for the smallest effect size of interest

We will begin with using values of d and N from the G*Power example  
Sample size N = 64  
cohen's d = 0.5  
number of iterations = 1000  
Use paramtest::run_test() to run our required number of iterations of the function
results go in the object "power_ttest"

```{r power_ttest, echo=TRUE}
power_ttest <-
  run_test(
    t_func,
    n.iter = 1000,
    output = 'data.frame',
    N = 64,
    d = .5
  )
```

### report the proportion of "sig" values that are TRUE

```{r results, echo=TRUE}
results(power_ttest) %>%
  summarise(power = mean(sig))
```

Compare the reported value with the value from G*Power (power = 0.8015)  
Note that it will be slightly different, but accuracy can be improved by increasing the number of iterations.  

### Use paramtest::grid_search() to explore scenarios

give the params argument a list of parameters we want to vary.  
lets try N=50, N=100 and N=150 (per group)

```{r scenario_N, echo=TRUE}
power_ttest_vary <-
  grid_search(
    t_func,
    params = list(N = c(50, 100, 150)),
    n.iter = 1000,
    output = 'data.frame',
    d = .5
  )
results(power_ttest_vary) %>%
  group_by(N.test) %>%
  summarise(power = mean(sig))
```
  
### Scenarios: 2 parameters varying

vary N and Cohen's d  
We will use values to cover the range shown in the G*Power example.  

```{r vary2, echo=TRUE}
power_ttest_vary2 <-
  grid_search(
    t_func,
    params = list(N = c(25, 50, 100, 200), d = c(.33, .5, .67)),
    n.iter = 1000,
    output = 'data.frame'
  )
power <- results(power_ttest_vary2) %>%
  group_by(N.test, d.test) %>%
  summarise(power = mean(sig))
print(power)
```
  
### Plot the power simulation

```{r plot2, echo=TRUE}
ggplot(power, aes(
  x = N.test,
  y = power,
  group = factor(d.test),
  colour = factor(d.test)
)) +
  geom_point() +
  geom_line() +
  ylim(c(0, 1)) +
  xlim(c(0, 200)) +
  labs(x = 'Sample Size', y = 'Power', colour = "Cohen's d") +
  theme_minimal()
```

Compare the plot with the G*Power plot.  They are roughly similar.  

Note: I have set n.iter = 1000 for each simulation to keep the run time very short.  Consider increasing n.iter to improve accuracy of results.

This simulation method can be generalised to a wide variety of statistical models.  Other packages may be preferred for specific situations.  

Reference: https://cran.r-project.org/web/packages/paramtest/vignettes/Simulating-Power.html
